Estimation of Stochastic Wilson and Cowan Neural Field Model using Expectation Maximisation Algorithm

### 2011-07-28 ltx/data_paper/DataPaper.tex
added algorithm

\usepackage{amsmath,amssymb,amsfonts} % Typical maths resource packages

\newif\ifpdf
\ifx\pdfoutput\undefined
\pdffalse % we are not running PDFLaTeX
\else
\pdfoutput=1 % we are running PDFLaTeX
\pdftrue
\fi

@@

meaning \eqref{eq:SecondDerivativexi} is negative and $\hat{\xi}$ represents a maximum. Equation \eqref{eq:SecondDerivativetheta} is negative where $\boldsymbol\Xi_4$ is positive definite requiring $\mathbf q(\mathbf x_t)$ to be of  rank $n_{\theta}$ \cite{Golub1996}. The matrix $\boldsymbol\Psi(\mathbf r')$ defined by \eqref{eq:DefPsi} is of rank $n_{\theta}$ as each row of the matrix is a set of equally spaced Gaussian basis functions with equal widths, consequently each row can be considered as a Riesz basis function for the subspaces they generate \cite{Ferrari2004} (note that multiplying by $\mathbf \Gamma^{-1}$ has no effect on the rank). However, $\mathbf q(\mathbf x_t)$ defined in \eqref{eq:QmatrixForSigmapoints} cannot be computed analytically as the inner product of a Gaussian and a sigmoid function cannot be solved analytically, making it difficult to calculate the rank of $\mathbf q(\mathbf x_t)$ . Therefore, the rank of $\mathbf q(\mathbf x_t)$ at each iteration of the algorithm should be examined. From \eqref{eq:SecondDerivativenoise} and \eqref{eq:SecondDerivativedisturbance}, observation noise and disturbance variance estimates clearly represent maxima of the $\mathcal{Q}$-function.

>>>>

\usepackage{amsmath,amssymb,amsfonts} % Typical maths resource packages
% Algorithms
\usepackage{float}
\floatstyle{ruled}
\newfloat{algorithm}{htbp}{loa}
\floatname{algorithm}{Algorithm}
\newif\ifpdf
\ifx\pdfoutput\undefined
\pdffalse % we are not running PDFLaTeX
\else
\pdfoutput=1 % we are running PDFLaTeX
\pdftrue
\fi

@@

meaning \eqref{eq:SecondDerivativexi} is negative and $\hat{\xi}$ represents a maximum. Equation \eqref{eq:SecondDerivativetheta} is negative where $\boldsymbol\Xi_4$ is positive definite requiring $\mathbf q(\mathbf x_t)$ to be of rank $n_{\theta}$ \cite{Golub1996}.  However, $\mathbf q(\mathbf x_t)$ defined in \eqref{eq:QmatrixForSigmapoints} cannot be computed analytically as the inner product of a Gaussian and a sigmoid function cannot be found analytically \cite{MacKay1992}, making it difficult to calculate the rank of $\mathbf q(\mathbf x_t)$. Therefore, the rank of $\mathbf q(\mathbf x_t)$ at each iteration of the algorithm should be examined numerically. In the experiments of this work the problem of rank deficiency was never encountered. From \eqref{eq:SecondDerivativenoise} and \eqref{eq:SecondDerivativedisturbance}, the observation noise and disturbance variance estimates clearly represent maxima of the $\mathcal{Q}$-function. The joint state and parameter estimation algorithm is summarised in Algorithm~\ref{alg:EM-EMAlgorithm}.
\begin{algorithm}\footnotesize
\caption{Joint state and parameter estimation for the IDE model.}
\label{alg:EM-EMAlgorithm}
\begin{enumerate}
\item  Initialise:
Generate a bounded, random sequence of dimension $n_x$ and use as the initial state sequence estimate.
\item Approximate $\Xi$-variables
	\boldsymbol\Xi_{0}&=\sum_{t=0}^{T-1}\left(\mathbf P_{t+1}^b+\mathbf{\hat x}_{t+1}^b\mathbf{\hat x}_{t+1}^{b\top}\right)\nonumber \\
\boldsymbol\Xi_{1}&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_{t+1}^{b\top}\right)\nonumber\\
\boldsymbol\Xi_{2}&=\sum_{t=0}^{T-1}\left(\mathbf P_t^b+\mathbf{\hat x}_t^b\mathbf{\hat x}_t^{b\top}\right)\nonumber\\
\boldsymbol\Xi_{3}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t+1}^{b\top}\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf{\hat x}_t^b)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf M_{t+1}^b \tilde{\boldsymbol\Sigma}_e^{-1} \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'\right]  \nonumber \\	
\boldsymbol\Xi_{4}&\approx\sum_{t=0}^{T-1}\left[\mathbf q^\top(\mathbf{\hat x}_t^b)\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf{\hat x}_t^b)+\sum_{i,j=1}^{n_x}[\mathbf{P}_t^b]_{ij}{\boldsymbol\Lambda}_t^{(i)\top}\tilde{\boldsymbol\Sigma}_e^{-1}{\boldsymbol\Lambda}_t^{(j)}\right]\nonumber\\
 \boldsymbol\Xi_{5}&\approx\sum_{t=0}^{T-1}\left[ \mathbf{\hat x}_{t}^{b\top}\tilde{\boldsymbol\Sigma}_e^{-1}\mathbf q(\mathbf{\hat x}_t^b)+\int_\Omega\boldsymbol \phi^\top(\mathbf r') \mathbf P_t^b \tilde{\boldsymbol\Sigma}_e^{-1}  \boldsymbol{\Psi}(\mathbf{r}') f'(\boldsymbol \phi^\top(\mathbf r')\mathbf {\hat x}_t^b) d\mathbf{r}'\right] \nonumber
\item Calculate the parameter estimates
 \hat{\xi}&=\frac{\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_3^\top}{\mathrm{tr}\left\lbrace \boldsymbol\Xi_2 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace-\boldsymbol\Xi_5\boldsymbol\Xi_4^{-1}\boldsymbol\Xi_5^{\top}} \nonumber\\
 \hat{\boldsymbol \theta}&=\boldsymbol\Xi_4^{-1}\left(\boldsymbol\Xi_3^\top-\hat{\xi}\boldsymbol\Xi_5^\top \right)\nonumber\\
\hat{\sigma}_{\epsilon}^2&=\frac{1}{(T-1)n_y}\sum_{t=0}^{T-1}\mathrm{tr}\left\lbrace (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1}) (\mathbf y_{t+1}-\mathbf C\mathbf{\hat{x}}_{t+1})^\top+\mathbf C \mathbf P_{t+1}\mathbf C^\top \right\rbrace \nonumber\\
2\hat{\xi}\mathrm{tr}\left\lbrace \boldsymbol\Xi_1 \tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace +\hat{\xi}^2\mathrm{tr} \left\lbrace\boldsymbol\Xi_2\tilde{\boldsymbol\Sigma}_e^{-1}\right\rbrace\right.\nonumber \\ &\left.-2\boldsymbol\Xi_3\hat{\boldsymbol\theta}+\hat{\boldsymbol\theta}^\top\boldsymbol \Xi_4\hat{\boldsymbol\theta}+2\hat{\xi}\boldsymbol\Xi_5 \hat{\boldsymbol\theta}\right] \nonumber
\item E step: Execute Algorithm~2.1 of chapter 2, and perform the additional step
\mathbf M_{t +1}^{b}=\mathbf S_t\mathbf P_{t+1}^{b} \nonumber
and compute the $\Xi$-variables using step 2.
\item M step: execute step 3.
\item Calculate stopping criterion.
\end{enumerate}
\end{algorithm}
